Astro 507; Problem Set 3

### 2022-02-23 pset3/solutions.tex
3b
3c

\question{3a. \textbf{Fugacity}}
\answer{
    The general goal here is to find the number density based on the central temperature and density and then solve for the Fermi-Dirac integral and we can then invert that to find the fugacity. To start, we know that
    \begin{equation}
        \rho = m_e n_e + m_{\rm H} n_{\rm H} + m_{\rm He} n_{\rm He}
    \end{equation}
    We can rewrite this (and neglect the electron term) as
    \begin{equation}
        \rho = n_{\rm H} \qty(m_{\rm H} + m_{\rm He} \frac{n_{\rm He}}{n_{\rm H}})
    \end{equation}
    Now we can plug in the masses of hydrogen and helium as well as their relative abundance (given as 0.1 in the problem description).
    \begin{equation}
        n_{\rm H} = \frac{\rho_c}{1.4 m_p}
    \end{equation}
    Now we just relate this to the electron number density simply as
    \begin{equation}
        n = \qty(2 \frac{n_{\rm He}}{n_{\rm H}} + 1) n_{\rm H}
    \end{equation}
    And combining the two gives an expression for the number density in terms of the central density
    \begin{equation}
        n = \frac{1.2 \rho_c}{1.4 m_p}
    \end{equation}
    The other part that we need is the de-Broglie wavelength which depends on the central temperature as
    \begin{equation}
        \lambda = \frac{h}{\sqrt{2 \pi m k_B T_c}}
    \end{equation}
    Now let's use this with the equation for the number density from Lecture 10 and solve for the Fermi-Dirac integral (plugging in values in the final equation).
    \begin{align}
        n &= \frac{2(2s + 1)}{n \pi^{1/2} \lambda^3} F_{1/2}(z) \\
        F_{1/2}(z) &= \frac{n \pi^{1/2} \lambda^3}{2(2s + 1)} \\
        F_{1/2}(z) &= \frac{\pi^{1/2}}{2(2s + 1)} \frac{1.2 \rho_c}{1.4 m_p} \qty(\frac{h}{\sqrt{2 \pi m k_B T_c}})^3 \\
        F_{1/2}(z) &= 2.08
    \end{align}
    I took this value and inverted it using the approximate analytic expressions from the paper (Aymerich-Humet+1981) to find the fugacity.
    \begin{equation}
        \boxed{ z = 1.65 }
    \end{equation}
}

>>>>

\question{3a. \textbf{Fugacity}}
\answer{
    The general goal here is to find the number density based on the central temperature and density and then solve for the Fermi-Dirac integral and we can then invert that to find the fugacity. To start, we know that
    \begin{equation}
        \rho = m_e n_e + m_{\rm H} n_{\rm H} + m_{\rm He} n_{\rm He}
    \end{equation}
    We can rewrite this (and neglect the electron term) as
    \begin{equation}
        \rho = n_{\rm H} \qty(m_{\rm H} + m_{\rm He} \frac{n_{\rm He}}{n_{\rm H}})
    \end{equation}
    Now we can plug in the masses of hydrogen and helium as well as their relative abundance (given as 0.1 in the problem description).
    \begin{equation}
        n_{\rm H} = \frac{\rho_c}{1.4 m_p}
    \end{equation}
    Now we just relate this to the electron number density simply as
    \begin{equation}
        n = \qty(2 \frac{n_{\rm He}}{n_{\rm H}} + 1) n_{\rm H}
    \end{equation}
    And combining the two gives an expression for the number density in terms of the central density
    \begin{equation}
        n = \frac{1.2 \rho_c}{1.4 m_p}
    \end{equation}
    The other part that we need is the de-Broglie wavelength which depends on the central temperature as
    \begin{equation}
        \lambda = \frac{h}{\sqrt{2 \pi m_e k_B T_c}}
    \end{equation}
    Now let's use this with the equation for the number density from Lecture 10 and solve for the Fermi-Dirac integral (plugging in values in the final equation).
    \begin{align}
        n &= \frac{2(2s + 1)}{n \pi^{1/2} \lambda^3} F_{1/2}(z) \\
        F_{1/2}(z) &= \frac{n \pi^{1/2} \lambda^3}{2(2s + 1)} \\
        F_{1/2}(z) &= \frac{\pi^{1/2}}{2(2s + 1)} \frac{1.2 \rho_c}{1.4 m_p} \qty(\frac{h}{\sqrt{2 \pi m_e k_B T_c}})^3 \\
        F_{1/2}(z) &= 2.08
    \end{align}
    I took this value and inverted it using the approximate analytic expressions from the paper (Aymerich-Humet+1981) to find the fugacity.
    \begin{equation}
        \boxed{ z = 1.65 }
    \end{equation}
}

\question{3b. \textbf{Pressure}}
\answer{
    In order to calculate the pressure we need to apply the same pressure equation as 2a
    \begin{equation}
        P = \frac{4(2s + 1)}{3 \pi^{1/2}} k_B T \lambda^{-3} F_{3/2}(z)
    \end{equation}
    In this specific case, $s = 1/2$, $T = 6 \times 10^6 \unit{K}$, $\lambda$ is calculated in the same way as part a and we found that $z = 1.65$ in part a. This gives the pressure as
    \begin{equation}
        \boxed{ P = 1.92 \times 10^{16} \unit{Pa} }
    \end{equation}
}

\question{3c. \textbf{Relativistic?}}
\answer{
    In lecture we found that electrons are relativistic when their fermi momentum is such that
    \begin{equation}
        p_{\rm F} \sim m_e c
    \end{equation}
    We showed that in terms of density this was equivalent to
    \begin{equation}
        \rho_{\rm rel} \sim 2 \times 10^6 \unit{g}{cm^{-3}} \cdot \frac{\mu_e}{2 m_p}
    \end{equation}
    The last term with $\mu_e$ is going to be approximately of order unity and so, since the central density is only $325 \unit{g}{cm^{-3}}$ we have that the electrons \textbf{are} relativistic
    \begin{equation}
        \boxed{ \rho_c \ll \rho_{\rm rel} \implies \text{non-relativistic} }
    \end{equation}
}
